---
title: "DDSAnalytics Attrition Analysis"
author: "Karen, Laurence, Sam"
date: "April 4, 2019"
output: 
        html_document:
        keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- install packages hidden-->
```{r echo=FALSE, include=FALSE}
    install.packages("caret", repos="http://cran.us.r-project.org")
    install.packages('tidyverse', repos="http://cran.us.r-project.org")
    install.packages("rsample",repos="http://cran.us.r-project.org")
    install.packages('broom', repos="http://cran.us.r-project.org")
    install.packages('corrplot', repos="http://cran.us.r-project.org")
    install.packages('ISLR', repos="http://cran.us.r-project.org")
    install.packages("caret", repos="http://cran.us.r-project.org")
    install.packages("klaR",repos="http://cran.us.r-project.org")
    install.packages("e1071",repos="http://cran.us.r-project.org")
    #install.packages('leaps', repos="http://cran.us.r-project.org")
    install.packages('MASS', repos="http://cran.us.r-project.org")
    
    #reference libraries
    library(tidyverse)
    library(rsample) 
    library(klaR)
    library(broom)
    library(corrplot)
    library(ISLR)
    library(readxl)
    library(caret)
    library(e1071)
    #library(leaps)
    library(MASS)
    library(ggplot2)
    
logit <- function (ps)
  ## define the logit function.
{
  log(ps/(1-ps))
}
```
## Introduction
According to the Society for Human Resource Management's 2016 *Human Capital Benchmarking Report*, the average cost-per-hire for companies is $4,129 and the average time to fill a vacant position is 42 days. In addtittion to the recruitment and training costs high attrition, can also face increased opertional and reduced productivity as they lose trained and experienced employees.

Controlling attrition is without a doubt an area of opportunity for all companies, and one that can greatly impact long term success. The following report details the findings and recommendations of DDSAnalytics after having reviewed ACME Corporations turnover and HR data set. 

Source: https://www.shrm.org/hr-today/trends-and-forecasting/research-and-surveys/Documents/2016-Human-Capital-Report.pdf

## The Data

The Dataset was comprised of 1470 observations each with 35 recorded variables. The data tracked a variety of factors possibly related to attrition. Based on the recorded summary data the attrition rate for the company is estimated to be 16.1% and the estimated average tenure is approximately 7 years, though it must be noted that based on a maximum recorded employee number of 2068 we can infer that there are at least 598 missing records. As such, the data will be treated as a sample of the overall population of company employees. While it is our goal to identify correlations between variables and create a model that accurately predicts attrition, because the dataset is comprised of observational data, no inferences regarding causation should be drawn from the results.


```{r ReadFile}
#Read in the CaseStudy2-data.xlsx and set to Attrition data frame
attrition <- read_excel("C:/git/CaseStudy2-master/CaseStudy2-data.xlsx")
#attrition<-read_excel("/Users/ksomes/Downloads/CaseStudy2_2/CaseStudy2-data.xlsx")
```

## Excluded Variables Initial Analysis
Upon the initial overview of the dataset it was determined that the variables "Standard Hours", "Over 18" and "Employee Count" would be excluded from exploratory data analysis as the value for those variables accross all observations is the same and would therefore impact the response variable for each observation with the same weight. The variable "Employee Number" was excluded from exploratory analysis because it is merely an identification variable and should have no effect on the response variable.

Exclusion of those variables resulted in a dataset with the following 30 explanatory variables. These variables would be used to determine their correlation to the response variable Attrition and to create a model for predicting Attrition. 

|Nominal                    |Ordinal                    |Continous                        |
|---------------------------|---------------------------|---------------------------------| 
|BusinessTravel             |EnvironmentSatisfaction    |Age                              |
|Department                 |JobInvolvement             |DailyRate                        |
|Education                  |JobLevel                   |DistanceFromHome                 |
|EducationField             |JobSatisfaction            |HourlyRate                       |
|JobRole                    |PerformanceRating          |MonthlyIncome                    |
|MaritalStatus              |RelationshipSatisfaction   |MonthlyRate                      |
|Gendter                    |StockOptionLevel           |NumCompaniesWorked               |
|Overtime                   |WorkLifeBalance            |PercentSalaryHike                |
|                           |                           |TotalWorkingYears                |
|                           |                           |TrainingTimesLastYear            |
|                           |                           |YearsAtCompany                   |
|                           |                           |YearsInCurrentRole               |
|                           |                           |YearsSinceLastPromotion          |
|                           |                           |YearsWithCurrManager             |


```{r}
# Drop Employee Count, Standard Hours, Over 18, and Employee number
attrition <- attrition[,-c(9,10,22,27),drop=FALSE]
```

## Exploration of Probabilities: Naive Bayes 
```{r echo=FALSE, include=FALSE}
#new data frame of categorical variables
dfcat <- attrition

#Exclude DailyRate, EmployeeCount, EmployeeNumber, HourlyRate, StandardHours, Over18, PerformanceRating, MonthlyRate
dfcat$DailyRate <- NULL
dfcat$EmployeeCount <- NULL
dfcat$EmployeeNumber <- NULL
dfcat$HourlyRate <- NULL
dfcat$StandardHours <- NULL
dfcat$Over18 <- NULL
dfcat$PerformanceRating <- NULL
dfcat$MonthlyRate <- NULL

#Categorize Variables
#Categorize Age
dfcat$Agecat<-cut(dfcat$Age, seq(18,60,10), right=FALSE,labels=c("Under_27","28-37","38-47",">48"))
dfcat$Age <- NULL

#Categorize Distance from Home
dfcat$DistanceFromHome<-cut(dfcat$DistanceFromHome, seq(0,30,10), right=FALSE, labels=c("Under_10","10_to_20","Over_20"))

#Categorize Education
dfcat$Education[dfcat$Education==1] = "Below_College"
dfcat$Education[dfcat$Education==2] = "College"
dfcat$Education[dfcat$Education==3] = "Bachelor"
dfcat$Education[dfcat$Education==4] = "Master"
dfcat$Education[dfcat$Education==5] = "Doctor"

#Categorize JobInvolement
dfcat$JobInvolvement[dfcat$JobInvolvement == 1]="Low"
dfcat$JobInvolvement[dfcat$JobInvolvement == 2]="Medium"
dfcat$JobInvolvement[dfcat$JobInvolvement == 3]="High"
dfcat$JobInvolvement[dfcat$JobInvolvement == 4]="Very_High"

#Categorize JobSatisfaction
dfcat$JobSatisfaction[dfcat$JobSatisfaction==1]="Low"
dfcat$JobSatisfaction[dfcat$JobSatisfaction==2]="Medium"
dfcat$JobSatisfaction[dfcat$JobSatisfaction==3]="High"
dfcat$JobSatisfaction[dfcat$JobSatisfaction==4]="Very_High"


#Categorize Monthly Income
dfcat$MonthlyIncome<-cut(dfcat$MonthlyIncome, seq(0,20000,5000), right=FALSE, labels=c("Under_5,000","Under_10,000","Under_15,000","Under_20,000"))

#Categorize NumCompaniesWorked
dfcat$NumCompaniesWorked<-cut(dfcat$NumCompaniesWorked, seq(0,10,5), right=FALSE, labels=c("Less_than_5","Greater_than_5"))

#Categorize PercentSalaryHike
dfcat$PercentSalaryHike<-cut(dfcat$PercentSalaryHike, seq(10,25,5), right=FALSE, labels=c("Less_than_15%","Less_than_20%","Greater_than_20%"))

#Catelgorize RelationshipSatisfaction
dfcat$RelationshipSatisfaction[dfcat$RelationshipSatisfaction == 1]="Low"
dfcat$RelationshipSatisfaction[dfcat$RelationshipSatisfaction == 2]="Medium"
dfcat$RelationshipSatisfaction[dfcat$RelationshipSatisfaction == 3]="High"
dfcat$RelationshipSatisfaction[dfcat$RelationshipSatisfaction == 4]="Very_High"

#Categorize EnvironmentSatisfaction
dfcat$EnvironmentSatisfaction[dfcat$EnvironmentSatisfaction==1]="Low"
dfcat$EnvironmentSatisfaction[dfcat$EnvironmentSatisfaction==2]="Medium"
dfcat$EnvironmentSatisfaction[dfcat$EnvironmentSatisfaction==3]="High"
dfcat$EnvironmentSatisfaction[dfcat$EnvironmentSatisfaction==4]="Very_High"

#Categorize StockOptionLevel
dfcat$StockOptionLevel[dfcat$StockOptionLevel==0]="None"
dfcat$StockOptionLevel[dfcat$StockOptionLevel==1]="Low"
dfcat$StockOptionLevel[dfcat$StockOptionLevel==2]="Medium"
dfcat$StockOptionLevel[dfcat$StockOptionLevel==3]="High"

#Categorize TotalWorkingYears
dfcat$TotalWorkingYears<-cut(dfcat$TotalWorkingYears, seq(0,40,10), right=FALSE, labels=c("Less_than_10","11_to_20","21_to_30","Gt_30"))

#Categorize TrainingTimesLastYear
dfcat$TrainingTimesLastYear<-cut(dfcat$TrainingTimesLastYear, seq(0,6,2), right=FALSE, labels=c("Less_than_two","Two-Four","Four-Six"))

#Categorize Work Life Balance
dfcat$WorkLifeBalance[dfcat$WorkLifeBalance==1]="Bad"
dfcat$WorkLifeBalance[dfcat$WorkLifeBalance==2]="Good"
dfcat$WorkLifeBalance[dfcat$WorkLifeBalance==3]="Better"
dfcat$WorkLifeBalance[dfcat$WorkLifeBalance==4]="Best"

#Categorize YearsAtCompany
dfcat$YearsAtCompany<-cut(dfcat$YearsAtCompany, seq(0,40,5), right=FALSE, labels=c("Five_or_Less","Six_to_Ten","Eleven_to_Fifteen","Sixteen_to_Twenty","TwentyOne_to_TwentyFive","TwentySix_to_Thirty","ThirtyOne_to_ThirtyFive","Greater_than_ThrityFive"))

#Categorize YearsInCurrentRole
dfcat$YearsInCurrentRole<-cut(dfcat$YearsInCurrentRole, seq(0,20,5), right=FALSE, labels=c("Five_or_Less","Six_to_Ten","Eleven_to_Fifteen","Sixteen_to_Twenty"))

#Categorize YearsSinceLastPromotion
dfcat$YearsSinceLastPromotion<-cut(dfcat$YearsSinceLastPromotion, seq(0,15,5), right=FALSE, labels=c("Five_or_Less","Six_to_Ten","Eleven_to_Fifteen"))

#Categorize YearsWithCurrManager
dfcat$YearsWithCurrManager<-cut(dfcat$YearsWithCurrManager, seq(0,20,5), right=FALSE, labels=c("Five_or_Less","Six_to_Ten","Eleven_to_Fifteen", "Sixteen_to_Twenty"))

```

```{r}
#Run Model and get probabilities
library(e1071)
NB_All<- naiveBayes(Attrition~. ,data=dfcat, scale=T, type="prob")
#list probabilities
NB_All$tables

```


```{r echo=FALSE, include=FALSE}
#Building the model
# Create training (70%) and test (30%) sets
set.seed(123)
#70 30 split
splitData <- initial_split(dfcat, prop = .7, strata = "Attrition")
train <- training(splitData)
test  <- testing(splitData)

# create response and feature data
features <- setdiff(names(train), "Attrition")
x <- train[, features]
y <- train$Attrition


# set up 5-fold cross validation procedure
#Tune model with Laplacian smoothing
tgrid <- data.frame(fL=c(0,0.5,1.0), usekernel = TRUE, adjust=c(0,0.5,1.0))

train_control <- trainControl(method = "cv", number = 5)

# train model
nbayes_model <- train(x = x,y = y,method = "nb",trControl = train_control,tuneGrid = tgrid)

```
## Prediction and cross validation of Naive Bayes Model
```{r }
#Prediction and cross validation
confusionMatrix(nbayes_model)


```
```{r echo=FALSE, include=FALSE}
#Assess accuracy
pred <- predict(nbayes_model, newdata = test)


```
```{r}
#Show graph of clustered yes no prediction responses

ggplot(test, aes(Attrition, pred, color = Attrition)) +
  geom_jitter(width = 0.2, height = 0.1, size=2) +
  labs(title="Confusion Matrix", 
       subtitle="Predicted Attrition vs. Actual Employee Attrition", 
       y="Predicted", 
       x="Truth",
       caption="NaiveBayes")

```


## Exploratory Data Analysis.  Interpreting the probabilities. 



```{r}


#Gender vs attrition.  Males tend to have a higher probability of attrition
Gender_prob <- subset(data.frame(NB_All$tables$Gender), Y=="Yes")
Gender_prob
barplot(Gender_prob$Freq, names.arg = Gender_prob$Gender, ylab="Attrition probability",main="Gender vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}

#Singles have a higher probability of attrition compared to married and divorced. 
MaritalStat_prob <- subset(data.frame(NB_All$tables$MaritalStatus),Y=="Yes")
MaritalStat_prob
barplot(MaritalStat_prob$Freq, names.arg = MaritalStat_prob$MaritalStatus, ylab="Attrition  Probability",main="Marital Status vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}

#Age.  Attrition probability is highest between 28 and 37 years of age.  The older the less likely to turn over. 
Age_prob <- subset(data.frame(NB_All$tables$Agecat),Y=="Yes")
Age_prob
barplot(Age_prob$Freq, names.arg = Age_prob$Agecat, ylab="Attrition  Probability",main="Age vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}

#Age.  Working Years vs Attrition.  The more work experience the less likely to turn over. 
WorkYears_prob <- subset(data.frame(NB_All$tables$TotalWorkingYears),Y=="Yes")
WorkYears_prob
barplot(WorkYears_prob$Freq, names.arg = WorkYears_prob$TotalWorkingYears, ylab="Attrition  Probability",main="Working Years vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}

#Income vs Attrition.  The lower the income level the higher the probability of attrition
MontlyI_prob <- subset(data.frame(NB_All$tables$MonthlyIncome),Y=="Yes")
MontlyI_prob
barplot(MontlyI_prob$Freq, names.arg = MontlyI_prob$MonthlyIncome, ylab="Attrition  Probability",main="Monthly Income vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}

#Salary hike vs Attrition.  The lower the hike the higher the probability of attrition
SalaryH_prob <- subset(data.frame(NB_All$tables$PercentSalaryHike),Y=="Yes")
SalaryH_prob
barplot(SalaryH_prob$Freq, names.arg = SalaryH_prob$PercentSalaryHike, ylab="Attrition  Probability",main="Percent Salary Hike vs Attrition", cex.names=0.7,col="red",las=1)

```

```{r}

#stock Option Levels vs attrition.  The lower the options the higher the attrition
StockO_prob <- subset(data.frame(NB_All$tables$StockOptionLevel),Y=="Yes")
StockO_prob
barplot(StockO_prob$Freq, names.arg = StockO_prob$StockOptionLevel, ylab="Attrition Probability",main="Stock Options vs Attrition", cex.names=0.7,col="red",las=1)

```
```{r}
#Attrition vs Opportunity to travel.  If you can travel, the opportunity to travel shows a lower probability of attrition.
BusTrav_prob <- subset(data.frame(NB_All$tables$BusinessTravel), Y=="Yes")
BusTrav_prob
barplot(BusTrav_prob$Freq, names.arg = BusTrav_prob$BusinessTravel, ylab="Attrition  Probability",main="Opportunity to Travel vs Attrition", cex.names=0.7,col="red",las=1)
```

```{r}

#Naive Bayes: Attrition vs Education. The higher level of education, the lower probability of attrition
Education_prob <- subset(data.frame(NB_All$tables$Education), Y=="Yes")
Education_prob
barplot(Education_prob$Freq ,names.arg=Education_prob$Education,ylab="Probability of Attrition",main="Education vs Attrition", cex.names=0.7,col="red",las=2)

```

```{r}
#Department vs probability of attrition.  Technical field tend to have a higher probability of attrition.  Human resources have a low probability of attrition in comparison.
Depart_prob <- subset(data.frame(NB_All$tables$Department),Y=="Yes")
Depart_prob
barplot(Depart_prob$Freq, names.arg = Depart_prob$Department, ylab="Attrition  Probability",main="Department vs Attrition", cex.names=0.4,col="red",las=2)

```

```{r}
#Job Involvment.  High degrees of job involvment such as those in technical field have a highe probability of attrition. However, if you are a decision maker such as a Chief Data Scientist very high involvment have low probabilities of attrition.
JobI_prob <- subset(data.frame(NB_All$tables$JobInvolvement), Y=="Yes")
JobI_prob
barplot(JobI_prob$Freq, names.arg = JobI_prob$JobInvolvement, ylab="Attrition  Probability",main="Job Involvement vs Attrition", cex.names=0.6,col="red",las=1)
```
```{r}
#Work life balance. On surveys 'Better' may be and indicator of employee unhappiness. A mark of 'Better' has a high probability of attrition
LifeBalance_prob <- subset(data.frame(NB_All$tables$WorkLifeBalance),Y=="Yes")
LifeBalance_prob
barplot(LifeBalance_prob$Freq, names.arg = LifeBalance_prob$WorkLifeBalance, ylab="Attrition  Probability",main="Work Life Balance vs Attrition", cex.names=0.6,col="red",las=1)

```


## Evaluating the correlations.  Logistric Regression.


```{r}
# Create a new binary variable for attrition
attrition$AttrV<-ifelse(attrition$Attrition=="Yes",1,0)
#create subset of Continous Variables
attritionInt <- attrition[,c(1,4,6,11,17:19,21,25,26,28:31,32)]
#create correlation matix for Continous Variables
corAttInt <- cor(attritionInt)
#plot correlation matrix
corrplot(corAttInt, order='AOE', method='circle')
#create smaller matrix focused on AttrV variable / Plot Matrix
corAttrVInt <- corAttInt[c(11,15),]
corrplot(corAttrVInt, method='circle',addCoef.col ='black', addCoefasPercent = TRUE )
```

## Preliminary Correlation Tests for for Ordinal Variables


```{r}
#create subset of Ordinal Variables
attritionOrd <- attrition[,c(9,12,13,15,22:24,27,32)]
#create correlation matix for Ordinal Variables
corAttOrd <- cor(attritionOrd)
#plot correlation matrix
corrplot(corAttOrd,order='AOE', method='circle',addCoef.col ='black', addCoefasPercent = TRUE )
#create smaller matrix focused on AttrV variable / Plot Matrix
corAttrVOrd <- corAttOrd[c(3,9),]
corrplot(corAttrVOrd, method='circle',addCoef.col ='black', addCoefasPercent = TRUE )
```

## Preliminary t-tests for significance of Ordinal and Continous Variables.


```{r echo=TRUE, results='hide', include=FALSE}
t.test(Age ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(DailyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(DistanceFromHome ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(Education ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(EnvironmentSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(HourlyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobInvolvement ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobLevel ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(MonthlyIncome ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(MonthlyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(NumCompaniesWorked ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(PercentSalaryHike ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(RelationshipSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(StockOptionLevel ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(TotalWorkingYears ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(TrainingTimesLastYear ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(WorkLifeBalance ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsAtCompany ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsInCurrentRole ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsSinceLastPromotion ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsWithCurrManager ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
```

|Category                   |t-value|	p-value |
|-------------------------  |:-----:|:-------:|
|TotalWorkingYears          |6.65   |< .0001  |
|JobLevel                   |6.57   |< .0001  |
|YearsInCurrentRole         |6.23   |< .0001  |
|MonthlyIncome              |6.20   |< .0001  |
|Age                        |6.18   |< .0001  |
|YearsWithCurrManager       |6.06   |< .0001  |
|StockOptionLevel           |5.30   |< .0001  |
|YearsAtCompany             |5.20   |< .0001  |
|JobInvolvement             |5.02   |< .0001  |
|JobSatisfaction            |3.99   |< .0001  |
|EnvironmentSatisfaction    |3.98   |< .0001  |
|DistanceFromHome           |2.99   |.0027    |
|WorkLifeBalance            |2.45   |.0142    |
|TrainingTimesLastYear      |2.28   |.0225    |
|DailyRate                  |2.17   |.0299    |
|RelationshipSatisfaction   |1.76   |.0787    |
|NumCompaniesWorked         |1.67   |.0955    |
|YearsSinceLastPromotion    |1.27   |.2058    |
|Education                  |1.20   |.2293    |
|MonthlyRate                |.58    |.5611    |
|PercentSalaryHike          |.52    |.6056    |
|HourlyRate                 |.26    |.7931    |


## Preliminary Correlation Tests for for Nominal Variables

```{r echo=TRUE, include=FALSE}
#create subsets for Catergorical and Dichonomous Variables
attritionCat <- attrition[,c(3,5,7,8,14,15,32)]
attritionDic <- attrition[,c(10,20,32)]
#create ANOVA tables
aovTravel = aov(attrition$AttrV ~ attrition$BusinessTravel)
aovDept= aov(attrition$AttrV ~ attrition$Department)
aovEd = aov(attrition$AttrV ~ attrition$Education) 
aovEdFld = aov(attrition$AttrV ~ attrition$EducationField)
aovRole = aov(attrition$AttrV ~ attrition$JobRole)
aovMStat = aov(attrition$AttrV ~ attrition$MaritalStatus)
aovGen = aov(attrition$AttrV ~ attrition$Gender)
aovOT = aov(attrition$AttrV ~ attrition$OverTime)
summary(aovTravel)
summary(aovDept)
summary(aovEd)
summary(aovEdFld)
summary(aovRole)
summary(aovMStat)
summary(aovGen)
summary(aovOT)
```

|Category         |**r**  |F-value  |*p*-value  |
|-----------------|:-----:|:-------:|:---------:|
|Over Time        |n/a    |94.66    |< .0001    |
|MaritalStatus    |.176   |23.78    |< .0001    |
|BusinessTravel   |.126   |12.27    |< .0001    |
|JobRole          |.243   |11.38    |< .0001    |
|Department       |.084   |5.43     |.0045      |
|Educationfield   |.105   |3.23     |.0066      |
|Education        |.044   |1.45     |.2290      |
|Gender           |n/a    |1.28     |.2591      |

```{r}
#create  model dataset with chosen variables
#attritionMdl <- attrition[,c(1,3,5,6,8,9,13:17,20,24,25,28,29,32),drop=FALSE]
#attritionMdl$AttrV <- factor(attritionMdl$AttrV, levels = c(0, 1))
#create full dataset with scrubbed data 
attritionFull <- as.data.frame(attrition)
attritionFull$AttrV <-ifelse(attrition$Attrition=="Yes",1,0)
```


## Creating a Training Model

```{r}
#create training data sample size at 80% of Dataset set.seed to ensure same rows are chosen
smpSize = floor(.80*nrow(attritionFull))
set.seed(123)
trainSmp = sample(seq_len(nrow(attritionFull)),size = smpSize)
trainAttr =attritionFull[trainSmp,] 
testAttr=attritionFull[-trainSmp,]
trainSmpF = sample(seq_len(nrow(attritionFull)),size = smpSize)
trainAttrF =attritionFull[trainSmp,-2] 
testAttrF=attritionFull[-trainSmp,]
```

## Model Selection
The response variable of interest, Attrition, is binary. In order to determine what leads to attrition, we decided to use a logistic regression model, a classification regression method for binary outputs. We assume that there is a linear, additive relationship between the predictors and the natural log odds of attrition rate.


## Stepwise Regression Variable Selection    
In order to determine which factors have a higher association with attrition rate, two systematic approaches were used: Forward and Backward stepwise variable selection.  In both cases, the three variables selected are Marital Status, Overtime, and Total Working Years. The importance of these factors will be further assessed in the building of a logistic regression classification model.    

```{r echo=FALSE, include=FALSE}
install.packages("leaps", repos="http://cran.us.r-project.org") 
library(leaps)
```
```{r}
# Fit the full model
StpFor<- regsubsets(AttrV~.,data=trainAttrF,method="forward",nbest=1, nvmax=3)
topFor<-coef(StpFor,3)
topFor
StpBack <- regsubsets(AttrV~.,data=trainAttrF,method="backward",nbest=1, nvmax=3)
topBack<-coef(StpBack,3)
topBack
    
```    

## Train Logistic Regression Model

```{r}
#create logitFull
logitFull <- glm(AttrV ~ YearsAtCompany + BusinessTravel +	Department + EducationField + Gender +	JobRole + MaritalStatus + Age + HourlyRate + DailyRate + MonthlyRate + MonthlyIncome + DistanceFromHome + Education + EnvironmentSatisfaction + JobInvolvement + JobLevel + JobSatisfaction + NumCompaniesWorked + PercentSalaryHike + PerformanceRating + RelationshipSatisfaction + StockOptionLevel + TotalWorkingYears + TrainingTimesLastYear + WorkLifeBalance + YearsInCurrentRole + YearsSinceLastPromotion + YearsWithCurrManager + OverTime, family = binomial, data = trainAttrF)
summary(logitFull)
anova(logitFull, test="Chisq")
cor(trainAttr[,c(31,30,29,28)])
#remove insignificant terms based on p-values and chi-squared tests & correlated terms
logitMod<-glm(AttrV~OverTime+MaritalStatus+TotalWorkingYears+BusinessTravel+DistanceFromHome
              +EnvironmentSatisfaction+JobInvolvement+JobRole+JobSatisfaction+MaritalStatus
              +NumCompaniesWorked+TotalWorkingYears+OverTime+
              RelationshipSatisfaction,data=trainAttrF, family=binomial(link="logit"))
summary(logitMod)
anova(logitMod, test="Chisq")
```

## Variable Selection
Variables were selected in the model by assessing whether adding the term had an impact on the model through analysis of variance tests and t-tests against the parameter value. Terms were removed from the model if the coefficient had a large p-value. For variables with similar p-values, the analysis of variance test was run on the model to determine which term explained the least amount of variation in attrition rate. As a result of this experimentation, the final model includes 11 terms: Business Travel, Distance From Home, Environment Satisfaction, Job Involvement, Job Role, Job Satisfaction, Marital Status, Number of Companies Worked, Overtime, Relationship Satisfaction, and Years With Current Manager. In the context of this model based on analysis of variance, the three most important factors that attribute to attrition are Overtime, Marital Status, and Total Working Years, which are the same factors selected by the stepwise algorithms. 

#Visualizing Key Factors
```{r}
by.t<-attritionFull %>% group_by(TotalWorkingYears) %>% summarize(attt=sum(AttrV))
prop.t<-by.t$attt/sum(by.t$attt)
plot(by.t$TotalWorkingYears,logit(prop.t), xlab="Working Years", ylab="Log Odds of Attrition Rate")
```

## Testing the Model
```{r}
#testing Data
pred <- predict(logitMod, newdata = testAttr, type = "response")
pred
# Recode factors
y_pred_num <- ifelse(pred > 0.5, 1, 0)
y_pred <- factor(y_pred_num, levels=c(0, 1))
y_act <- testAttr$AttrV
# Accuracy
mean(y_pred == y_act)  
predF <- predict(logitFull, newdata = testAttrF, type = "response")
predF
# Recode factors
y_pred_numF <- ifelse(predF > 0.5, 1, 0)
y_predF <- factor(y_pred_numF, levels=c(0, 1))
y_actF <- testAttrF$AttrV
# Accuracy
mean(y_predF == y_actF)  
```    

The logistic regression model based on 11 variables had a 88% accurate classification rate against the test data. While overtime eligibility, marriage status, and years in the workforce are key indicators, there are many factors that lead to attrition in a company.

## Other Job Role Trends
While attrition is a great concern for companies, it is also important for management to understand the working culture that they create. Below we will discuss some key employee-related findings.    

```{r}
ggplot(attritionFull, aes(x=reorder(JobRole, MonthlyRate, FUN = median), y=MonthlyRate, color=JobRole)) + geom_boxplot()+theme(
        axis.text.x=element_blank(),axis.title.x = element_blank())
```    
The monthly rate appears to be distributed evenly across all job roles; no job role has a significantly higher monthly rate than the others.

```{r}
ggplot(attritionFull, aes(x=reorder(JobRole, Age, FUN=median),y=Age, color=JobRole))+geom_boxplot()+theme(
        axis.text.x=element_blank(), axis.title.x = element_blank())
```    
The distribution of ages differs across job roles. Sales Representatives tend to be the youngest, while managers and directors tend to be older. Healthcare representatives tend to be older than manufacturing directors, which is interesting considering it is not a traditional management role (management roles tend to be associated with experience and therefore age).     


```{r}
ggplot(attritionFull, aes(x=MonthlyIncome,y=Age))+geom_point()+geom_smooth(method=lm)
```
Not many young employees have high monthly incomes, but low monthly incomes are associated with a wide range of ages.


```{r}
ggplot(attritionFull, aes(x=YearsSinceLastPromotion, y=Age))+geom_point()+geom_smooth(method=lm)
```    
Older employees tend to be promoted less often, while younger employees tend to more actively "climb the ladder". 






