---
title: "DDSAnalytics Attrition Analysis"
author: "Karen, Laurence, Sam"
date: "April 4, 2019"
output: 
        html_document:
        keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- install packages hidden-->
```{r echo=FALSE, include=FALSE}

    install.packages('tidyverse', repos="http://cran.us.r-project.org")
    install.packages('broom', repos="http://cran.us.r-project.org")
    install.packages('corrplot', repos="http://cran.us.r-project.org")
    install.packages('ISLR', repos="http://cran.us.r-project.org")
    #install.packages('caret', repos="http://cran.us.r-project.org")
    #install.packages('leaps', repos="http://cran.us.r-project.org")
    install.packages('MASS', repos="http://cran.us.r-project.org")
    
    
    library(tidyverse)
    library(broom)
    library(corrplot)
    library(ISLR)
    library(readxl)
    #library(caret)
    #library(leaps)
    library(MASS)
    

```
## Introduction
According to the Society for Human Resource Management's 2016 *Human Capital Benchmarking Report*, the average cost-per-hire for companies is $4,129 and the average time to fill a vacant position is 42 days. In addtittion to recruitment and training costs high attrition can often mean that companies who lose experienced employees face the additional opertional costs of lost productivity. 
Controling attrition is without a doubt an area of opportunity for all companies and one that can greatly impact long term success. The following report details the findings and recommendations of DDSAnalytics after having reviewed ACME Corporations turnover and HR data set. 

Source: https://www.shrm.org/hr-today/trends-and-forecasting/research-and-surveys/Documents/2016-Human-Capital-Report.pdf

## The Data

The Dataset was comprised of 1470 observations each with 35 recorded variables. The data tracked a variety of factors possibly related to attrition. Based on the recorded summary data the attrition rate for the company is estimated to be 16.1%, though it must be noted that based on a maximum recorded employee number of 2068 we can infer that there are at least 598 missing records. As such, the data will be treated as a sample of the overall population of company employees. While it is our goal to identify correlation and create a model that accurately predicts attrition, because the dataset is comprised of observational data, no inferences regarding causation should be drawn from the results.


```{r ReadFile}

#Read in the CaseStudy2-data.xlsx and set to Attrition data frame
attrition <- read_excel("C:/CaseStudy2/CaseStudy2-data.xlsx")


```

## Excluded Variables Initial Analysis
Upon the initial overview of the dataset it was determined that the variables "Standard Hours", "Over 18" and "Employee Count" would be excluded from exploratory data analysis as the value for those variables accross all observations is the same and would therefore impact the response variable for each observation with the same weight. The variable "Employee Number" was excluded from exploratory analysis because it is merely an identification variable and should have no effect on the response variable.

Exclusion of those variables resulted in a dataset with the following 31 variables. 

|Dichotomous  |Nominal            |Ordinal                  |Interval/Ratio           |
|-------------|-----------------------|-------------------------|-------------------------| 
|Attrition    |BusinessTravel         |EnvironmentSatisfaction  |Age                      |
|Gender       |Department             |JobInvolvement           |DailyRate                |
|OverTime     |Education              |JobLevel                 |DistanceFromHome         |
|             |EducationField         |JobSatisfaction          |HourlyRate               |
|             |JobRole                |PerformanceRating        |MonthlyIncome            |
|             |MaritalStatus          |RelationshipSatisfaction |MonthlyRate              |
|             |                       |StockOptionLevel         |NumCompaniesWorked       |
|             |                       |WorkLifeBalance          |PercentSalaryHike        |
|             |                       |                         |TotalWorkingYears        |
|             |                       |                         |TrainingTimesLastYear    |
|             |                       |                         |YearsAtCompany           |
|             |                       |                         |YearsInCurrentRole       |
|             |                       |                         |YearsSinceLastPromotion  |
|             |                       |                         |YearsWithCurrManager     |


```{r}
# Drop Employee Count, Standard Hours, Over 18, and Employee number
attrition <- attrition[,-c(9,10,22,27),drop=FALSE]

```
## Preliminary Correlation Tests for Interval and Ratio Variables

```{r}
# Create a new binary variable for attrition
attrition$AttrV <- ifelse(attrition$Attrition == 'Yes', attrition$AttrV <- 1, ifelse(attrition$Attrition == 'No', attrition$AttrV <- 0, attrition$Attrition <- NA))

#create subset of Interval/Ratio Variables
attritionInt <- attrition[,c(1,4,6,11,17:19,21,25,26,28:31,32)]


#create correlation matix for Interval/Ratio Variables
corAttInt <- cor(attritionInt)

#plot correlation matrix
corrplot(corAttInt, order='AOE', method='circle')

#create smaller matrix focused on AttrV variable / Plot Matrix
corAttrVInt <- corAttInt[c(11,15),]
corrplot(corAttrVInt, method='circle',addCoef.col ='black', addCoefasPercent = TRUE )

```

## Preliminary Correlation Tests for for Ordinal Variables


```{r}
#create subset of Ordinal Variables
attritionOrd <- attrition[,c(9,12,13,15,22:24,27,32)]

#create correlation matix for Ordinal Variables
corAttOrd <- cor(attritionOrd)

#plot correlation matrix
corrplot(corAttOrd,order='AOE', method='circle',addCoef.col ='black', addCoefasPercent = TRUE )


#create smaller matrix focused on AttrV variable / Plot Matrix
corAttrVOrd <- corAttOrd[c(3,9),]
corrplot(corAttrVOrd, method='circle',addCoef.col ='black', addCoefasPercent = TRUE )

```

## Preliminary t-tests of Interval, Ratio, and Ordinal Variables for significance

```{r echo=TRUE, include=FALSE}

t.test(Age ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(DailyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(DistanceFromHome ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(Education ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(EnvironmentSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(HourlyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobInvolvement ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobLevel ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(JobSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(MonthlyIncome ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(MonthlyRate ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(NumCompaniesWorked ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(PercentSalaryHike ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(RelationshipSatisfaction ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(StockOptionLevel ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(TotalWorkingYears ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(TrainingTimesLastYear ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(WorkLifeBalance ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsAtCompany ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsInCurrentRole ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsSinceLastPromotion ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')
t.test(YearsWithCurrManager ~ AttrV, data=attrition, var.equal=T, alternative='two.sided')

```

|Category	                |t-value|	p-value |
|-------------------------|:-----:|:-------:|
|TotalWorkingYears        |6.65   |< .0001  |
|JobLevel	                |6.57	  |< .0001  |
|YearsInCurrentRole	      |6.23	  |< .0001  |
|MonthlyIncome	          |6.20	  |< .0001  |
|Age                      |6.18   |< .0001  |
|YearsWithCurrManager     |6.06   |< .0001  |
|StockOptionLevel         |5.30	  |< .0001  |
|YearsAtCompany	          |5.20	  |< .0001  |
|JobInvolvement 	        |5.02	  |< .0001  |
|JobSatisfaction	        |3.99	  |< .0001  |
|EnvironmentSatisfaction	|3.98	  |< .0001  |
|DistanceFromHome	        |2.99	  |.0027    |
|WorkLifeBalance	        |2.45	  |.0142    |
|TrainingTimesLastYear    |2.28   |.0225    |
|DailyRate	              |2.17	  |.0299    |
|RelationshipSatisfaction	|1.76	  |.0787    |
|NumCompaniesWorked	      |1.67	  |.0955    |
|YearsSinceLastPromotion  |1.27	  |.2058    |
|Education	              |1.20	  |.2293    |
|MonthlyRate	            |.58	  |.5611    |
|PercentSalaryHike	      |.52	  |.6056    |
|HourlyRate               |.26	  |.7931    |


## Preliminary Correlation Tests for for Nominal and Dichtomous Variables

```{r echo=TRUE, include=FALSE}

#create subsets for Catergorical and Dichonomous Variables
attritionCat <- attrition[,c(3,5,7,8,14,15,32)]
attritionDic <- attrition[,c(10,20,32)]

#create ANOVA tables
aovTravel = aov(attrition$AttrV ~ attrition$BusinessTravel)
aovDept= aov(attrition$AttrV ~ attrition$Department)
aovEd = aov(attrition$AttrV ~ attrition$Education) 
aovEdFld = aov(attrition$AttrV ~ attrition$EducationField)
aovRole = aov(attrition$AttrV ~ attrition$JobRole)
aovMStat = aov(attrition$AttrV ~ attrition$MaritalStatus)
aovGen = aov(attrition$AttrV ~ attrition$Gender)
aovOT = aov(attrition$AttrV ~ attrition$OverTime)

summary(aovTravel)
summary(aovDept)
summary(aovEd)
summary(aovEdFld)
summary(aovRole)
summary(aovMStat)
summary(aovGen)
summary(aovOT)
```

|Category         |**r**  |F-value  |*p*-value  |
|-----------------|:-----:|:-------:|:---------:|
|Over Time        |n/a    |94.66    |< .0001    |
|MaritalStatus    |.176   |23.78    |< .0001    |
|BusinessTravel   |.126   |12.27    |< .0001    |
|JobRole          |.243   |11.38    |< .0001    |
|Department       |.084   |5.43     |.0045      |
|Educationfield   |.105   |3.23     |.0066      |
|Education        |.044   |1.45     |.2290      |
|Gender           |n/a    |1.28     |.2591      |

## Variable Selection and Model Choice
After examining the data it was decided to reduce the variables to be tested down to a set of 16 variables: Age, Business Travel, Department, DistanceFromHome, EducationField, EnvironmentSatisfaction, JobLevel, JobRole, JobSatisfaction, Marital Status, MonthlyIncome, OverTime, StockOptionLevel, TotalWorkingYears, YearsAtCompany, YearsInCurrentRole

Because the response variable, Attrition, does not lend itself well to linear regression, it was decided that a logistic regression would be a better model.

```{r}
#create  model dataset with chosen variables
attritionMdl <- attrition[,c(1,3,5,6,8,9,13:17,20,24,25,28,29,32),drop=FALSE]
attritionMdl$AttrV <- factor(attritionMdl$AttrV, levels = c(0, 1))

#create full dataset with scrubbed data 
attritionFull <- as.data.frame(attrition)
attritionFull$AttrV <- factor(attrition$AttrV, levels = c(0, 1))

```

## Stepwise Regression Variable Selection


```{r}
# Fit the full model
#attritionStp <- glm(AttrV ~ ., family = binomial, data = attrition)

#Stepwise regression model
#attritionStp <- stepAIC(attritionFull, direction = "both", 
#                      trace = FALSE)
#summary(attritionStep)
    
```


## Creating a Training Model

```{r}
#create training data sample size at 80% of Dataset set.seed to ensure same rows are chosen
smpSize = floor(.80*nrow(attritionMdl))
set.seed(123)

trainSmp = sample(seq_len(nrow(attritionMdl)),size = smpSize)
trainAttr =attritionMdl[trainSmp,] 
testAttr=attritionMdl[-trainSmp,]

trainSmpF = sample(seq_len(nrow(attritionFull)),size = smpSize)
trainAttrF =attritionFull[trainSmp,] 
testAttrF=attritionFull[-trainSmp,]

#create logitmodel 
logitMod <- glm(AttrV ~ Age + BusinessTravel + Department + DistanceFromHome + EducationField + EnvironmentSatisfaction + JobLevel + JobRole + JobSatisfaction + MaritalStatus + MonthlyIncome + OverTime + StockOptionLevel + TotalWorkingYears + YearsAtCompany + YearsInCurrentRole, family =binomial, data = trainAttr)

#create logitFull
logitFull <- glm(AttrV ~ YearsAtCompany + BusinessTravel +	Department + EducationField + Gender +	JobRole + MaritalStatus + Age + HourlyRate + DailyRate + MonthlyRate + MonthlyIncome + DistanceFromHome + Education + EnvironmentSatisfaction + JobInvolvement + JobLevel + JobSatisfaction + NumCompaniesWorked + PercentSalaryHike + PerformanceRating + RelationshipSatisfaction + StockOptionLevel + TotalWorkingYears + TrainingTimesLastYear + WorkLifeBalance + YearsInCurrentRole + YearsSinceLastPromotion + YearsWithCurrManager + OverTime, family = binomial, data = trainAttrF)

summary(logitMod)

summary(logitFull)

# compare models
anova(logitFull,logitMod)



```

## Testing the Model
```{r}

#testing Data
pred <- predict(logitMod, newdata = testAttr, type = "response")
pred

# Recode factors
y_pred_num <- ifelse(pred > 0.5, 1, 0)
y_pred <- factor(y_pred_num, levels=c(0, 1))
y_act <- testAttr$AttrV

# Accuracy
mean(y_pred == y_act)  

predF <- predict(logitFull, newdata = testAttrF, type = "response")
predF

# Recode factors
y_pred_numF <- ifelse(predF > 0.5, 1, 0)
y_predF <- factor(y_pred_numF, levels=c(0, 1))
y_actF <- testAttrF$AttrV

# Accuracy
mean(y_predF == y_actF)  
```


